Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-09-10 intro.tex
wrote about running coupling

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Introduction}
The standard model of particle physics (SM) is our current best working
theory that describes the phenomenon we observe, for  example,
 at collider experiments.
The large hadron collider (LHC) is a high energy proton-proton
collider experiment that allows us to study the fundamental particles
laid out in the standard model.

The interaction of these fundamental particles are described
by scattering amplitudes. The standard model allows us to
calculate these amplitudes, or matrix elements, from first
principles. A key challenge in the past few decades has been
to compute these matrix elements for increasing number of
final state particles and for higher orders in perturbative
expansations in fixed order calculations. Although there
has been tremendous progress in this sector a complication
that naturally arises is that the evaluation of these matrix
elements becomes increasingly time consuming due to their
algebraic complexity.

@@

\label{chapter:intro}
\section{The Standard Model of particle physics}
    The Standard Model of particle physics (SM) is our current
    best working theory to describe all known elementary particles
    as well as three of the four fundamental forces. Developed
    predominantly in the latter half of the 20th century, it is
    one of the most well tested theories we have in science today.
    Some highlights include the highly precise predictions of the
    anomalous magnetic moment of the electron, agreeing with experimental
    measurements to more than 10 significant figures \cite{Aoyama:2017uqe}.
    and the discovery of the Higgs boson in 2012 by the ATLAS \cite{ATLAS:2012yve}
    and CMS experiments \cite{CMS:2012qbp} at the Large Hadron Collider
    (LHC) which was theorised decades prior.

@@

    The particle content of the SM is summarised in Figure \ref{fig:SM_particles}
    showing the quarks, leptons and bosons along with their masses, charges
    and spin.
    \begin{figure}
        \includegraphics[width=\linewidth]{introduction/sm_particles.pdf}
        \caption{Particle content of the SM split into quarks, leptons,
        gauge bosons and scalar bosons. The columns for the fermions depict
        the three different generations. Masses, electric charge, and spin
        are given for each particle. The yellow contours indicate the coupling
        of bosons to fermions, illustrating the forces experienced by the
        fermions inside the contour. Figure reference \cite{SM_figure}.}
        \label{fig:SM_particles}
    \end{figure}
    The interactions of these fields are governed by the SM Lagrangian
    \footnote{technically Lagrangian density but we use the terms
    Lagrangian and Lagrangian density interchangeably}
    which can be written as
    \begin{equation}\label{eqn:L_SM}
        \mathcal{L}_{\mathrm{SM}} = \mathcal{L}_{\mathrm{gauge}} + \mathcal{L}_{\mathrm{fermion}} + \mathcal{L}_{\mathrm{Higgs}} + \mathcal{L}_{\mathrm{Yukawa}} + \mathcal{L}_{\mathrm{GF}} + \mathcal{L}_{\mathrm{ghost}} \, ,
    \end{equation}
    where $\mathcal{L}_{\mathrm{gauge}}$ describes the gauge fields,
    $\mathcal{L}_{\mathrm{fermion}}$ describes how fermions interact with
    gauge fields as well as their kinetic terms,
    $\mathcal{L}_{\mathrm{Higgs}}$ describes the Higgs field,
    $\mathcal{L}_{\mathrm{Yukawa}}$ describes the interaction between the Higgs
    field and fermions,
    $\mathcal{L}_{\mathrm{GF}}$ is a gauge fixing term,
    and $\mathcal{L}_{\mathrm{ghost}}$ is a ghost term.
    The last two terms are required to remove unphysical degrees of freedom
    when gauge fixing the theory.
    All terms in the SM Lagrangian are invariant under local transformations
    of the gauge group (\ref{eqn:SM_gauge}).

@@

\section{Introduction to quantum chromodynamics}
    Quantum chronodynamics (QCD) is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-abelian gauge theory with gauge group
    $\mathrm{SU}(3)_{C}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{f}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{\mathrm{s}}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{\mathrm{s}} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The gauge coupling of the group $g_{\mathrm{s}}$
    is a dimensional free parameter of the theory, usually
    written as
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{\mathrm{s}} = \dfrac{g_{\mathrm{s}}^{2}}{4\pi} \, ,
    \end{equation}
    for reasons which will become clear when we discuss
    scattering amplitudes.

@@

\section{Scattering amplitudes and cross-sections}
    In collider experiments we typically collide
    two beams, consisting of bundles of energetic
    particles, and analyse the resulting products
    of these collisions. To model collisions in
    a collider experiment, we look at the specific
    case of two particles colliding into $n$ particles.
    To mathematically describe these
    scattering processes, scattering amplitudes are used.
    For an initial-state
    $|i\rangle$ and final-state $|f\rangle$
    the scattering amplitude can be written as
    \begin{equation}\label{eqn:scattering}
        \langle f | S | i \rangle \, ,
    \end{equation}
    where the scattering matrix, or $S$-matrix
    can be decomposed into an identity matrix
    and a transfer matrix $\mathcal{T}$,
    \begin{equation}\label{eqn:S_matrix}
        S = \mathbbm{1} + \mathrm{i}\mathcal{T} \, .
    \end{equation}
    The $S$-matrix encodes all the information
    about how the initial-state will evolve
    into the final-state over time. By writing
    the $S$-matrix in this way, all the interactions
    are separated into the transfer matrix $\mathcal{T}$,
    as the identity matrix describes the free theory.
    By imposing a momentum conserving constraint
    on the $S$-matrix, the matrix element,
    $\mathcal{M}$, is defined in the following expression
    \begin{equation}\label{eqn:matrix_element}
        \langle f | \mathcal{T} | i \rangle = (2\pi)^{4} \delta^{4}\left(p_{a} + p_{b} - \sum_{f=1}^{n} p_{f}^{\mu}\right) \mathcal{M} \, .
    \end{equation}
    where we take $p_{a}$ and $p_{b}$ to be
    the momenta of the two colliding
    particles in the initial-state, and $p_{f}$ to be the
    momenta of the $n$-particle final states.
    The transition probability of this scattering process
    is the modulus squared of the scattering amplitude
    \begin{equation}\label{eqn:S_prob}
        P = |\langle f | S | i \rangle|^{2} \propto |\langle f | \mathcal{M} | i \rangle 
    \end{equation}
    where $|\langle f | \mathcal{M} | i \rangle|^{2}  \equiv |\mathcal{M}|^{2}$
    is the matrix element squared.

@@

    Since detectors at collider experiments
    have limited resolution, single scattering
    events are not measured. Instead we consider
    the cross-section, $\sigma$, which is a property
    of the particles being scattered, and is independant
    of the way the experiment is carried out.
    The cross-section directly relates the
    matrix elements which we calculate in our QFT,
    to measurable observables at experiments.
    In practice, it is more useful to consider the
    differential cross-section, $\mathrm{d}\sigma$
    where the cross-section can be differential
    in quantities such as energies and angles.
    This is because the final-state particles
    will be produced with a range of, for example,
    energies and momenta so the cross-section would
    be vanishing for an exact momentum configuration.

>>>>

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Introduction}
\label{chapter:intro}
The standard model of particle physics (SM) is our current best working
theory that describes the phenomenon we observe, for  example,
 at collider experiments.
The large hadron collider (LHC) is a high energy proton-proton
collider experiment that allows us to study the fundamental particles
laid out in the standard model.

The interaction of these fundamental particles are described
by scattering amplitudes. The standard model allows us to
calculate these amplitudes, or matrix elements, from first
principles. A key challenge in the past few decades has been
to compute these matrix elements for increasing number of
final state particles and for higher orders in perturbative
expansions in fixed order calculations. Although there
has been tremendous progress in this sector a complication
that naturally arises is that the evaluation of these matrix
elements becomes increasingly time-consuming due to their
algebraic complexity.

@@

\section{The Standard Model of particle physics}
    The Standard Model of particle physics (SM) is our current
    best working theory to describe all known elementary particles
    as well as three of the four fundamental forces. Developed
    predominantly in the latter half of the 20th century, it is
    one of the most well tested theories we have in science today.
    Some highlights include the highly precise predictions of the
    anomalous magnetic moment of the electron, agreeing with experimental
    measurements to more than 10 significant figures \cite{Aoyama:2017uqe}.
    and the discovery of the Higgs boson in 2012 by the ATLAS \cite{ATLAS:2012yve}
    and CMS experiments \cite{CMS:2012qbp} at the Large Hadron Collider
    (LHC) which was theorised decades prior.

@@

    The particle content of the SM is summarised in Figure \ref{fig:SM_particles}
    showing the quarks, leptons and bosons along with their masses, charges
    and spin.
    \begin{figure}
        % \includegraphics[width=\linewidth]{introduction/sm_particles.pdf}
        \caption{Particle content of the SM split into quarks, leptons,
        gauge bosons and scalar bosons. The columns for the fermions depict
        the three different generations. Masses, electric charge, and spin
        are given for each particle. The yellow contours indicate the coupling
        of bosons to fermions, illustrating the forces experienced by the
        fermions inside the contour. Figure reference \cite{SM_figure}.}
        \label{fig:SM_particles}
    \end{figure}
    The interactions of these fields are governed by the SM Lagrangian
    \footnote{technically Lagrangian density but we use the terms
    Lagrangian and Lagrangian density interchangeably}
    which can be written as
    \begin{equation}\label{eqn:L_SM}
        \mathcal{L}_{\mathrm{SM}} = \mathcal{L}_{\mathrm{gauge}} + \mathcal{L}_{\mathrm{fermion}} + \mathcal{L}_{\mathrm{Higgs}} + \mathcal{L}_{\mathrm{Yukawa}} + \mathcal{L}_{\mathrm{GF}} + \mathcal{L}_{\mathrm{ghost}} \, ,
    \end{equation}
    where $\mathcal{L}_{\mathrm{gauge}}$ describes the gauge fields,
    $\mathcal{L}_{\mathrm{fermion}}$ describes how fermions interact with
    gauge fields as well as their kinetic terms,
    $\mathcal{L}_{\mathrm{Higgs}}$ describes the Higgs field,
    $\mathcal{L}_{\mathrm{Yukawa}}$ describes the interaction between the Higgs
    field and fermions,
    $\mathcal{L}_{\mathrm{GF}}$ is a gauge fixing term,
    and $\mathcal{L}_{\mathrm{ghost}}$ is a ghost term.
    The last two terms are required to remove unphysical degrees of freedom
    when gauge fixing the theory.
    All terms in the SM Lagrangian are invariant under local transformations
    of the gauge group (\ref{eqn:SM_gauge}).

@@

\section{Introduction to quantum chromodynamics}
    Quantum chronodynamics (QCD) is the sector of the SM that
    describes the strong interaction. 
    QCD is a non-Abelian gauge theory with gauge group
    $\mathrm{SU}(3)_{C}$ where the charge is named colour.
    The gauge and fermion part of the QCD Lagrangian is
    \begin{equation}\label{eqn:L_QCD}
        \mathcal{L}_{\mathrm{QCD}} = -\dfrac{1}{4}F^{a}_{\mu\nu}F^{a, \, \mu\nu}
        + \sum_{f} \bar{\psi}_{i}^{f}(\mathrm{i}\slashed{D}_{ij} - \delta_{ij}m_{f})\psi_{j}^{f} \, ,
    \end{equation}
    where twice repeated indices are summed over.
    The fields $\psi_{i}^{f}$ ($\bar{\psi}_{i}^{f}$) are the fermions (antifermions),
    representing quarks (antiquarks) with flavours $f$: 
    up $u$, down $d$, strange $s$, charm $c$, top $t$, and bottom $b$,
    with masses $m_{f}$. They transform under the fundamental
    representation with indices $i, j \in \{1, 2, 3\}$, named colour indices.
    The gauge fields $A^{a}_{\mu}$, corresponding to gluons, 
    appear in the quark covariant derivative
    \begin{equation}\label{eqn:covariant_deriv}
        (D_{\mu})_{ij} = \delta_{ij}\partial_{\mu} - ig_{\mathrm{s}}T^{a}_{ij}A^{a}_{\mu} \, ,
    \end{equation}
    and the gauge field strength tensor
    \begin{equation}\label{eqn:F_munu}
        F^{a}_{\mu\nu} = \partial_{\mu}A^{a}_{\nu} - \partial_{\nu}A^{a}_{\mu} + g_{\mathrm{s}} f^{abc}A^{b}_{\mu}A^{c}_{\nu} \, .
    \end{equation}
    Gauge fields transform under the adjoint representation, meaning
    the adjoint indices $a,b,c \in \{1,...,8\}$. $T^{a}_{ij}$ are the
    group generators in a fundamental representation. In SU(3) it is
    common to write the group generators as
    \begin{equation}\label{eqn:group_generators}
        T^{a}_{ij} = \dfrac{1}{2}\lambda^{a}_{ij}
    \end{equation}
    where $\lambda^{a}_{ij}$ are the Gell-Mann matrices \cite{Gell-Mann:1962yej}.
    The generators of the group have to obey the Lie algebra
    \begin{equation}\label{eqn:lie_algebra}
        [T^{a}, T^{b}] = \mathrm{i}f^{abc}T^{c} \, ,
    \end{equation}
    where $f^{abc}$ are the structure constants of SU(3).
    The gauge coupling of the group $g_{\mathrm{s}}$
    is a dimensional free parameter of the theory, usually
    written as
    \begin{equation}\label{eqn:alpha_s}
        \alpha_{\mathrm{s}} = \dfrac{g_{\mathrm{s}}^{2}}{4\pi} \, .
    \end{equation}
    The coupling in QCD is dependent on the energy scale
    of the process, $\mu$. The behaviour of $\alpha_{\mathrm{s}}$
    is governed by the Callan-Symanzik \cite{Callan:1970yg,Symanzik:1970rt}
    $\beta$-function
    \begin{equation}\label{eqn:beta_fn}
        \mu^{2} \dfrac{\partial \alpha_{\mathrm{s}}(\mu^{2})}{\partial \mu^{2}} = \beta(\alpha_{\mathrm{s}}) \, ,
    \end{equation}
    where the $\beta$-function can be written as a perturbative
    expansion in $\alpha_{\mathrm{s}}$
    \begin{equation}\label{eqn:beta_series}
        -\beta(\alpha_{\mathrm{s}}) = \sum_{n=0}^{\infty}\dfrac{\alpha_{\mathrm{s}}^{n+2}}{(4\pi)^{n+1}} \beta_{n}\, .
    \end{equation}
    where the coefficients $\beta_{n}$ have been computed
    up to fifth order \cite{Baikov:2016tgj,Luthe:2017ttg}.
    The coefficients computed up to this point have been
    strictly positive, meaning that $\alpha_{\mathrm{s}}$
    decreases with increasing energy due to the minus sign
    in (\ref{eqn:beta_series}), a property known as
    asymptotic freedom \cite{Gross:1973id,Politzer:1973fx}.
    The solution to (\ref{eqn:beta_fn}) to first order
    is
    \begin{equation}\label{eqn:1l_alpha}
        \alpha_{\mathrm{s}}(\mu^{2}) = \dfrac{1}{\frac{\beta_{0}}{4\pi}\log(\frac{\mu^{2}}{\Lambda_{\mathrm{QCD}}^{2}})} \, ,
    \end{equation}
    where $\Lambda_{\mathrm{QCD}}$ is the QCD scale,
    the scale at which $\alpha_{\mathrm{s}}$ becomes
    large enough that perturbation theory breaks down.
    The running of the coupling constant has been
    observed experimentally as can be seen in Figure~\ref{fig:alpha_s_running}.
    \begin{figure}
        \includegraphics{alpha_s_running.pdf}
        \caption{The running of the strong coupling constant $\alpha_{\mathrm{s}}$,
        as determined by experiments, with QCD theory prediction in black.
        Figure reference \cite{Workman:2022ynf}.}
        \label{fig:alpha_s_running}
    \end{figure}

@@

\section{Scattering amplitudes and cross-sections}
    In collider experiments we typically collide
    two beams, consisting of bundles of energetic
    particles, and analyse the resulting products
    of these collisions. To model collisions in
    a collider experiment, we look at the specific
    case of two particles colliding into $n$ particles.
    To mathematically describe these
    scattering processes, scattering amplitudes are used.
    For an initial-state
    $|i\rangle$ and final-state $|f\rangle$
    the scattering amplitude can be written as
    \begin{equation}\label{eqn:scattering}
        \langle f | S | i \rangle \, ,
    \end{equation}
    where the scattering matrix, or $S$-matrix
    can be decomposed into an identity matrix
    and a transfer matrix $\mathcal{T}$,
    \begin{equation}\label{eqn:S_matrix}
        S = \mathbbm{1} + \mathrm{i}\mathcal{T} \, .
    \end{equation}
    The $S$-matrix encodes all the information
    about how the initial-state will evolve
    into the final-state over time. By writing
    the $S$-matrix in this way, all the interactions
    are separated into the transfer matrix $\mathcal{T}$,
    as the identity matrix describes the free theory.
    By imposing a momentum conserving constraint
    on the $S$-matrix, the matrix element,
    $\mathcal{M}$, is defined in the following expression
    \begin{equation}\label{eqn:matrix_element}
        \langle f | \mathcal{T} | i \rangle = (2\pi)^{4} \delta^{4}\left(p_{a} + p_{b} - \sum_{f=1}^{n} p_{f}^{\mu}\right) \mathcal{M} \, ,
    \end{equation}
    where we take $p_{a}$ and $p_{b}$ to be
    the momenta of the two colliding
    particles in the initial-state, and $p_{f}$ to be the
    momenta of the $n$-particle final states.
    The transition probability of this scattering process
    is the modulus squared of the scattering amplitude
    \begin{equation}\label{eqn:S_prob}
        P = |\langle f | S | i \rangle|^{2} \propto |\langle f | \mathcal{M} | i \rangle 
    \end{equation}
    where $|\langle f | \mathcal{M} | i \rangle|^{2}  \equiv |\mathcal{M}|^{2}$
    is the matrix element squared.

@@

    Since detectors at collider experiments
    have limited resolution, single scattering
    events are not measured. Instead we consider
    the cross-section, $\sigma$, which is a property
    of the particles being scattered, and is independent
    of the way the experiment is carried out.
    The cross-section directly relates the
    matrix elements which we calculate in our QFT,
    to measurable observables at experiments.
    In practice, it is more useful to consider the
    differential cross-section, $\mathrm{d}\sigma$
    where the cross-section can be differential
    in quantities such as energies and angles.
    This is because the final-state particles
    will be produced with a range of, for example,
    energies and momenta so the cross-section would
    be vanishing for an exact momentum configuration.
